Cheminformatics: The Computer Science of Chemical Discovery

### 2011-06-29 paper.tex
Added in Aarons updates

\title{Chemoinformatics: The Computer Science of Chemistry}
\numberofauthors{3}
\author{
\alignauthor
Ben Trovato\titlenote{Dr.~Trovato insisted his name be first.}\\
       \affaddr{Institute for Clarity in Documentation}\\
       \affaddr{1932 Wallamaloo Lane}\\
       \affaddr{Wallamaloo, New Zealand}\\
       \email{trovato@corporation.com}
% 2nd. author
\alignauthor
G.K.M. Tobin\titlenote{The secretary disavows
any knowledge of this author's actions.}\\
       \affaddr{Institute for Clarity in Documentation}\\
       \affaddr{P.O. Box 1212}\\
       \affaddr{Dublin, Ohio 43017-6221}\\
       \email{webmaster@marysville-ohio.com}
% 3rd author
\alignauthor
Rajarshi Guha\\
\affaddr{NIH Center for Translational Therapeutics}\\
\affaddr{9800 Medical Center Drive}\\
\affaddr{Rockvile, MD 20850}\\
\email{guhar@mail.nih.gov}
}

\additionalauthors{Additional authors: John Smith (The Th{\o}rv{\"a}ld Group,
email: {\texttt{jsmith@affiliation.org}}) and Julius P.~Kumquat
(The Kumquat Consortium, email: {\texttt{jpkumquat@consortium.net}}).}
\date{30 July 1999}

\maketitle
\begin{abstract}
  One of the most prominent success stories in all of science over the last decade has
  been the advance of bioinformatics: interdisciplinary collaboration between
  computer scientists and molecular biologists led to the sequencing the human genome
  (among many other accomplishments), because researchers connected biological
  concepts to the theory of string algorithms. However, despite this great
  success, few computer scientists are familiar with a related (and older!)
  discipline -- chemoinformatics, the use of computers to represent
  small molecules structures and analyze their properties based on aspects
  of their structures. Cheminformatics has a wide applicability
  including the discovery of novel drug candidates to optimizing
  certain physicochemical properties of small molecules. Until
  recently, the data and techniques of chemoinformatics have been closely guarded
  secrets of companies whose financial success depended on being the first to produce
  the new "miracle molecule". Only within the last decade -- and largely because of
  chemists volunteering their time for an Open Science "movement" -- do researchers
  now have access to freely available software packages and databases of tens of
  millions of chemicals. As a result, chemists now confront a variety
  of unsolved
  algorithmic problems that could not have been tackled a decade ago, but whose
  solutions are critical to research ranging from determining the behavior of small
  molecules in biological pathways, to finding a cure for malaria.
\end{abstract}

\category{H.4}{Information Systems Applications}{Miscellaneous}
\terms{Theory}
\keywords{cheminformatics, graphs, chemistry}

@@

Novel technologies in the life sciences churn out information at an ever
increasing rate -- public data stores such as at the European Bioinformatics
Institute (EBI) contain on the order of 10 PB of biological information.
However, while biological information has been available publicly for decades,
the same has not been true for chemical information until very recently. It was
not until 2004 that a large public small molecule structure repository (Pubchem)
became freely available. This was followed by other databases as discussed later
in this article. In the same vein, open source software and the publication of
algorithms have only recently become a main focus of the field
\cite{faulon2010}.

So why should we actually care about chemical information being made public --
and how does this relate to the field of computer science?

Both of these questions have relatively straightforward answers. We should care
about chemical information being made public since the drugs we take are the
result of intensive and costly research, and the more information is shared (a
process sometimes very hard to achieve in pharmaceutical companies) the more
easy it should be to develop novel treatments drugs. How chemical information
relates to computer science becomes clear when realizing the amount of data
available currently -- 74 million different molecules with annotations are
currently stored in public databases such as PubChem with a large amount of
annotations, and the design of data structures for chemicals, as well as
subsequent data mining of this information, are areas where expertise in
computer science is dearly needed. (For a recent, detailed introduction to the
field written for computer scientists see \cite{brown2009}).

Cheminformatics comprises different areas, which broadly speaking can be divided
into three fields: capturing data (using lab notebooks or potentially using
formats such as the Chemical Markup Language for publications); storing data
(designing database schemas, devising ontologies) and mining data (such as for
bioactivity prediction, which might involve algorithms based on graphs).
However, what is a major difference to the bioinformatics field is the type of
information being analyzed: While bioinformatics often deals with sequences, the
domain of cheminformatics is chemical structures. In case of the former,
information can very frequently be represented as one-dimensional strings, which
are relatively easy to handle computationally. On the other hand, chemical
structures may possess rings, branches as well as multiple valid representations
of even the same structure, potentially giving rise to ambiguities. Hence,
chemical structure is often thought of as being more difficult to standardize
than information from other domains. As an illustration, different
representations of the same chemical structure are shown in Figure 1. The top
left hand corner shows a structure a chemist would draw it -- every 'corner'
representing a carbon atom (which can also be represented as a C), every O
denoting oxygen, every N nitrogen, every H hydrogen and so on. This is the
standard format chemists use when exchanging information with their peers.
However, this is not directly amenable to data mining -- different atoms might
be collapsed into one node, sets of atoms might have the same relevant
properties assigned. Hence, the top right hand corner of Figure 1 shows an
annotated graph representation, taking into account that atoms, though being of
different elemental atom type, can have similar properties -- e.g. both oxygen
and nitrogen carry an 'On' label in this graph.

@@

On this graph, and employing large databases of thousands of molecules, graph
mining algorithms can be applied to discover patterns in large-scale datasets
and a large number of successful applications of this concept have been
published \cite{wegner2006,horst2009}. Still, this second representation is far
from being the end of the story. Molecules are 3D entities as represented in the
bottom left hand corner of the figure, and the precise 3D shape is relevant in
many situations (while not in others), but this is information that is lost when
only considering the graph of a molecule. To even extend this concept, in many
cases actually the surface properties of theq molecule are what appears to be
relevant in e.g. biological systems, as shown in the bottom right hand corner.
Hence, while graph mining on chemicals is certainly a promising avenue to
explore, one needs to keep in mind that not all molecular properties are encoded
in the graph (such as rotational states, or different charge states etc.) and
thus often other 'descriptors' need to be used to capture the essentials of a
chemical system \cite{bender2004}.

From the above it becomes clear that the nature of chemical problems is often a
complex mixture of combinatorial and continuous problems. If we oversimplify the
nature of chemical structures and their properties we can not describe the world
around us in sufficient details. On the other hand, if we try encoding all
details the complexity to collect, store, and mine data increases dramatically
(e.g. all possible 3D structures and their properties, all metabolites of a
structure in humans in different tissues). For the field of cheminformatics to
flourish, we need closer collaboration between chemists (or, more generally,
life scientists) and computer scientists – where the former needs to be able to
pose the problem in a way relevant for practical applications; and where the
latter is able to devise ways of capturing, storing, and analyzing chemical data
by bridging the gap between risk in generalizable encodings and complexity
(space and time). The goal is supporting a better chemical decision making by
reducing the risks by 1. storing and integrating data in maintainable ways, 2.
providing enough open standards and tools to allow software engineering concepts
to help bridging silos, and 3. mining the many chemical property spaces in a
time and space efficient way. Many aspects pertaining to those areas will be
discussed in the following sections.

@@

The motivation for this article is to highlight a variety of topics in
cheminformatics and present the computational and algorithmic problems
that are addressed in the field. The hope is that we will be able to
highlight areas where modern CS research could contribute to these
problems.

While cheminformatics has a broad range of applicability, ranging from
agrochemical research and drug discovery to the design of novel
materials, we have chosen to use drug discovery as the context of this
article. The remainder of this article is structured around the
concept of ``risk minimization'' in drug discovery -- how does one
minimize the chances of a small molecule failing as a drug candidate
and how do cheminformatics methods play a key role in this efforts.

@@

There is a considerable progress in open chemical and bioactivity databases like
ChEMBL and NCBI-PubChem, which are both freely available. Still, the integration
of chemical databases with each other remains a challenge, not only due to the
normalization problems in the chemical and bioactivity landscape, but also due
to the sheer data volume of them. As mentioned earlier, we have to be able to
search in complex data, for example multi-labelled molecular graphs, where
labels can be different tautomeric states, different protonation states, or many
variations of implicit chemical expert knowledge we need to capture in a
risk-bounded machine encoded way.

Chemical graphs can differ based on the manner in which the structures are
drawn. Checking whether two graphs represent the same chemical is an expensive
operation (the next section will discuss this in-depth), thus, a representation
of the structure that is unique and invariant with respect to how the structure
is drawn is needed for identification. An early effort in this direction was the
Simplified Molecular Input Line Specification (SMILES). However, several
different implementations became available. After years of research and various
commercial products, is there today not a single public standard for a
line-notation of molecular labelled graphs, which is available to a broad
audience (academic and commercial). The very same problem does also exist for
bioassay ontologies and the lack of options to normalize the property spaces of
chemical compounds, e.g. biological assays or clinical data outcome
(biobanking).

With each database using their own algorithm for molecular encoding, little
facilitated the automated exchange of data between different databases. As more
and more data became available online, structure-based identification became a
pressing need, and the IUPAC International Chemical Identifier (InChI) code was
developed. The InChI is a non-proprietary, structured textual identifier for
chemical entities [1]. InChI identifiers are not intended to be read and
understood by humans, but are useful for computational matching of chemical
entities. For rapid database lookups, the InChIKey is a hashed key for the
InChI, which has an invariant length of 14 characters. Figure 1 illustrates the
SMILES, InChI and InChIKey for lipoic acid.

Figure 1: The standard chemical graph illustration together with the SMILES,
InChI and InChIKey codes for lipoic acid. Note that the SMILES is much more
human-readable than the others.

The InChI is now widely used for matching identical chemical structures, but it
is still limited in a few scenarios. For example, it cannot differentiatebetween
certain types of stereoisomers: Figure 2 illustrates two stereoisomers for which
the generated InChI is the same.

Figure 2: The generated InChI is the same for the stereoisomers cisplatin and
transplatin.

InChI or other identity-mapping algorithms allow for exact searching (within the
limitations of the algorithms), but two other requirements for scientific
discovery based on chemical databases are substructure searching and similarity
searching. In substructure searching, the database is searched for a specified
wholly contained part of the structures in the database; in similarity
searching, the database is searched for structures similar to a provided search
structure. Chemical search packages are often implemented and optimised for a
given database technology, for example, the OrChem package is an open source
chemical search package for the Oracle database \cite{rijnbeek2009}.

Graph substructure matching is a variant of graph isomorphism detection,which is
well known to be computationally intractable [3]. To execute a graph isomorphism
search across a full chemical database of thousands or even millions of
structures is completely infeasible. For that reason, complex heuristics must be
used to drastically filter the candidate search space. Structure-based
fingerprints are commonly used for this purpose. Fingerprints encode
characteristic features of a given chemical structure in a boolean array, or
bitmap. The fingerprint is created by an algorithm which generates bitmap
patterns for each atom in the structure; then each atom and its nearest
neighbours, including the bonds between them; then each group of atoms connected
by paths of increasing lengths up to the maximum implemented for the algorithm
(commonly around eight). The bitmap patterns are added together into the final
fingerprint. If all bits in a query fingerprint are also present in the target
fingerprint of a stored database structure, only then is the structure subjected
to the computationally expensive subgraph matching algorithm. Bit operations are
very fast and independent of the number of atoms in a structure due to the fixed
length of the fingerprint.

Fingerprints also facilitate the rapid calculation of quantitative
structure-based similarity measures. An example measure of similarity is the
Tanimoto coefficient, calculated as the ratio $T(a,b) = \frac{c}{(a+b−c)}$,
where c is the count of bits on in the same position in both the two
fingerprints, a is the count of bits on in the first structure, and b is the
count of bits on in the second structure. The Tanimoto coefficient varies in the
range 0.0 -- 1.0, with a score of 1.0 indicating that the two structures are very
similar (i.e. their fingerprints are the same). Structure-based similarity is of
limited use in grouping of chemicals based on non-structural features, such as
shared bioactivity profiles. For this purpose, a large-scale and flexible
classification structure is needed for chemicals and their linked biological
information. One such answer to this need is provided in the form of ontologies.
An ontology is a formal specification of entities and their relationships in a
domain of interest. It is structured around an underlying logical representation
such as Description Logic [baaderdl2007]. The logic-based representation allows
automated reasoning to derive new knowledge (inferences) from the knowledge
encoded, making the representation both compact and immensely powerful.
Ontologies are already in widespread use across the biomedical domain. A
similarity measure based on the information encoded in ontologies is termed
‘semantic similarity, and such measures have already been applied to enhance
chemical classification in several problem areas [couto2010]. Many open
questionsstill remain as to the best way to combine the information encoded in
chemical structures with chemical ontologies [hastingsowled2010].

@@

Enumerating molecules is a combinatorial problem that has fascinated chemists,
computer scientists and mathematicians alike for more than a century. It is one
important risk-reduction element in still being able to encode closely-related
molecules and for keeping the space complexity of databases under (algorithmic)
control. Indeed, while trying to solve the problem of counting the isomers of
paraffin structures or counting substituted aromatic compounds, fundamental
principles in graph theory and combinatorics were developed by Cayley, Polya and
others. Even terms that are widely-used nowadays like graph and tree were
originally coined in a chemistry context (Chapters 1 and 8 in [1]). More than
four decades ago, Joshua Lederberg, Carl Djerassi, Edward Feigenbaum and others
from Stanford University started developing codes to explicitly enumerate
molecules. Their project named DENDRAL was funded by NASA with the goal of
developing an automated system to elucidate compounds from spectral data. While
studying this challenging problem, important concepts in computer science were
developed. DENDRAL is generally quoted in artificial intelligence textbooks as a
pioneer project that produced the first expert system.

Enumerating molecules is not only an interesting academic exercise but has
practical utility as well. The foremost application of enumeration is structure
elucidation of natural compounds such as metabolites, geochemicals, and
petrochemicals. Ideally, the wishful bench chemist collects experimental data
from an unknown compound, the data is fed into a code, and the resulting
structure is returned. Although that streamlined process may not yield a unique
solution, there are commercial software products such as MOLGEN
(http://molgen.de) that can, for instance, list all structures matching a given
molecular formula, or mass spectrometry (MS) data. Another important application
of enumeration is molecular design. Here the problem is to design compounds
(drugs, for instance) that optimize some physical, chemical, or biological
property or activity. Although less prolific and more recent than structure
elucidation, molecular design has introduced some novel solutions to molecular
enumeration. Finally, with the advent of combinatorial chemistry, molecular
enumeration has taken a central role as it allows computational chemists to
construct virtual libraries, test hypotheses, and provides guidance in the
design of optimal combinatorial experiments.

The major difficulty with enumeration is that the in silico representation of
molecules, the so-called molecular graphs, are labeled object (i.e. labeled
graphs), while the atoms of a molecule are of course not uniquely labeled. The
mathematical concept to tackle this problem is to consider orbits of labeled
molecular graphs under the operation of the symmetric group. When enumerating
molecular structures, one has thus to solve the so–called isomorphism problem
The good news is that isomorphism can be solved in polynomial time for molecular
graphs due to the fact that molecules belong to a restricted class of graphs
known as bounded valence graphs; however, that same restriction prevents
applying for molecular graph the polynomial delay enumeration results obtained
for general graphs (Chapter 8 in [1]). While the complexity of molecular graph
enumeration remains an open problem it has nonetheless been shown that molecules
can be sampled efficiently [2]. Sampling procedures based on the Metropolis or
the Genetic Algorithms have been developed and used to elucidate natural
compounds from NMR data.

The underlying computational complexity and potential intractability of
molecular enumeration may not be a bottleneck after all, as computational
chemists have developed and successfully used enumeration tools to generate
large chemical libraries like GDC-13 comprising almost a trillion chemical
structures (http://www.dcb-server.unibe.ch/groups/reymond/gdb/home.html). Yet,
the current enumeration software products do not generally produce stereoisomers
[3] and tautomers [4], these require specific enumeration procedures, which are
still the subject of investigations by the chemoinformatics community. GDB-13
along with other databases such as PubChem (NCBI/NIH) or ZINC (UCSF) are
valuable resources to search for new drugs. For instance the well known
ligand-based design approach is making use of these databases along with
structure similarity tools to discover new compounds binding to biological
targets (protein or nucleic acid). Compounds retrieved through ligand-based
design, however, must already be present in databases, which is a limitation of
the technique. To palliate that limitation algorithms have recently been
developed to enumerate all compounds in the chemical space matching circular [5]
or path fingerprints [6], these fingerprints being popular tools to measure
similarity. Enumerating molecules from fingerprints, or more generally
enumerating molecules from structure-activity molecular descriptors, is still
very much into development. At stake is the design of new chemicals or drugs
that have not yet been synthesized.

>>>>

\title{Cheminformatics: The Computer Science of Chemical Discovery}
\numberofauthors{9}
\author{
\alignauthor
Joerg Kurt Wegner\\
       \affaddr{Tibotec BVBA}\\
       \affaddr{Turnhoutseweg 30}\\
       \affaddr{2340 Beerse Turnhout, Belgium}\\
       \email{jwegner@its.jnj.com}
% 2nd. author
\alignauthor
Aaron Sterling\\
       \affaddr{Department of Computer Science}\\
       \affaddr{Iowa State University}\\
       \affaddr{Ames, Iowa, USA}\\
       \email{sterling@iastate.edu}
% 3rd author
\alignauthor
Rajarshi Guha\\
\affaddr{NIH Center for Translational Therapeutics}\\
\affaddr{9800 Medical Center Drive}\\
\affaddr{Rockvile, MD 20850}\\
\email{guhar@mail.nih.gov}
}

\additionalauthors{Additional authors:
Andreas Bender (University of Cambridge, email: {\texttt{bender.andreas@gmail.com}}),
Jean-Loup Faulon (University of Evry, email: {\texttt{Jean-Loup.Faulon@issb.genopole.fr}}),
Janna Hastings (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{janna.hastings@gmail.com}}),
Noel O'Boyle (University College, Cork, Ireland, email: {\texttt{baoilleach@gmail.com}}),
John Overington (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{jpo@ebi.ac.uk}}),
Herman Van Vlijmen (Tibotec, Beerse, Belgium, email: {\texttt{hvvlijme@its.jnj.com}}), and
Egon Willighagen (Karolinska Institutet, Stockholm, Sweden, email: {\texttt{baoilleach@gmail.com}})
.}
\date{25 June 2011}

\maketitle
\begin{abstract}
  One of the most prominent success stories in all of science over the last decade has
  been the advance of bioinformatics: interdisciplinary collaboration between
  computer scientists and molecular biologists led to the sequencing the human genome
  (among many other accomplishments), because researchers connected biological
  concepts to the theory of string algorithms. However, despite this great
  success, few computer scientists are familiar with a related (and older!)
  discipline---cheminformatics, the use of computers to represent
  small molecules structures and analyze their properties based on aspects
  of their structures. Cheminformatics has a wide applicability
  including the discovery of novel drug candidates to optimizing
  certain physicochemical properties of small molecules. Until
  recently, the data and techniques of chemoinformatics have been closely guarded
  secrets of companies whose financial success depended on being the first to produce
  the new ``miracle molecule.'' Only within the last decade---and largely because of
  chemists volunteering their time for an Open Science ``movement''---do researchers
  now have access to freely available software packages and databases of tens of
  millions of chemicals. As a result, chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago, but whose
  solutions are critical to research ranging from determining the behavior of small
  molecules in biological pathways, to finding a cure for malaria.
\end{abstract}

\category{H.4}{Information Systems Applications}{Miscellaneous}
\terms{
\textit{Tautomerism} - Tautomers are isomers of organic compounds that readily interconvert by a chemical reaction called tautomerization.
\textit{Protonation} - In chemistry, protonation is the addition of a proton (H+) to an atom, molecule, or ion.
}
\keywords{cheminformatics, graphs, chemistry, databases, knowledge management, open source, open data}

@@

Novel technologies in the life sciences churn out information at an ever
increasing rate---public data stores such as at the European Bioinformatics
Institute (EBI) contain on the order of 10 petabytes of biological information.
However, while biological information has been available publicly for decades,
the same has not been true for chemical information until very recently. It was
not until 2004 that a large public small molecule structure repository (Pubchem)
became freely available. This was followed by other databases, as discussed later
in this article. In the same vein, open source software and the publication of
algorithms have only recently become a main focus of the field
\cite{faulon2010}.

So why should we actually care about chemical information being made public---and how does this relate to the field of computer science?

We should care about chemical information being made public since the drugs we take are the
result of intensive and costly research, and the more information is shared (a
process sometimes very hard to achieve in pharmaceutical companies) the more
easy it should be to develop novel treatments drugs. How chemical information
relates to computer science becomes clear when realizing the amount of data
available currently---74 million different molecules are
currently stored in public databases (such as PubChem) with a large amount of
annotations, and the design of data structures for chemicals, as well as
subsequent data mining of this information, are areas where expertise in
computer science is dearly needed. (For a recent, detailed introduction to cheminformatics, written for computer scientists, see \cite{brown2009}).

Cheminformatics comprises different areas, which broadly speaking can be divided
into three fields: \emph{capturing data} (using lab notebooks or potentially using
formats such as the Chemical Markup Language for publications); \emph{storing data}
(designing database schemas, devising ontologies) and \emph{mining data} (such as for
bioactivity prediction, which might involve algorithms based on graphs).
However, a major difference from the bioinformatics field is the type of
information being analyzed: While bioinformatics often deals with sequences, the
domain of cheminformatics is chemical structures. In case of the former,
information can very frequently be represented as one-dimensional strings, which
are relatively easy to handle computationally. In the case of the latter, chemical
structures may possess rings, branches as well as multiple valid representations
of even the same structure, potentially giving rise to ambiguities. Hence,
chemical structure is often thought to be more difficult to standardize.

To illustrate the difficulty of standardizing chemical structures, different
representations of the same chemical structure are shown in Figure 1. The top
left hand corner shows a structure a chemist would draw it---every `corner'
representing a carbon atom (which can also be represented as a C), every O
denoting oxygen, every N nitrogen, every H hydrogen, and so on. This is the
standard format chemists use when exchanging information with their peers.
However, this format is not directly amenable to data mining---different atoms might
be collapsed into one node, sets of atoms might have the same relevant
properties assigned. Hence, the top right hand corner of Figure 1 shows an
annotated graph representation, taking into account that atoms, though being of
different elemental atom type, can have similar properties---\emph{e.g.}, both oxygen
and nitrogen carry an `On' label in this graph.

@@

Graph mining algorithms can be applied on this type of annotated graph (employing large databases of thousands of molecules),  to discover patterns in large-scale datasets.
Many successful applications of this concept have been
published \cite{wegner2006,horst2009}. Still, this ``annotated graph'' representation is far
from being the end of the story. Molecules are three-dimensional entities, as represented in the
bottom left hand corner of the figure, and the precise 3D shape is relevant in
many situations (while not in others).  Nevertheless, the 3D information is lost when
one only considers the graph of a molecule. To extend the concept of three-dimensionality further, in many
biological systems the \emph{surface properties} of the molecule are what appears to be
relevant, as shown in the bottom right hand corner.
Hence, while graph mining on chemicals is certainly a promising avenue to
explore, one needs to keep in mind that not all molecular properties are encoded
in the graph (for example, neither rotational states, nor different charge states are encoded) and
thus often other `descriptors' need to be used to capture the essentials of a
chemical system \cite{bender2004}.

From the above it becomes clear that the nature of chemical problems is often a
complex mixture of combinatorial and continuous problems. If we oversimplify the
nature of chemical structures and their properties, we can not describe the world
around us in sufficient detail. On the other hand, if we try to encode every
detail (\emph{e.g.}, all possible 3D structures and their properties, all metabolites of a
structure in humans in different tissues), the complexity to collect, store, and mine data increases dramatically. For the field of cheminformatics to
flourish, we need closer collaboration between chemists (or, more generally,
life scientists) and computer scientists---where the former need to be able to
pose their problems in a way relevant for practical applications; and where the
latter are able to devise ways of capturing, storing, and analyzing chemical data
by bridging the gap between risk in generalizable encodings and complexity
(space and time). The goals: to support better chemical decision making by 1. storing and integrating data in maintainable ways, 2.
providing enough open standards and tools so software engineering concepts
allow applications and data to be used across heterogeneous platforms, and 3. mining the many chemical property spaces in a time- and space-efficient way. We will discuss all three goals in the following sections.

@@

Our motivation for this article is to highlight a variety of topics in
cheminformatics and present the computational and algorithmic problems
that are addressed in the field. We hope to
highlight areas where modern CS research could contribute to these
problems.

While cheminformatics has a broad range of applicability, ranging from
agrochemical research and drug discovery to the design of novel
materials, we have chosen to use drug discovery as the context of this
article. The remainder of this article is structured around the
concept of ``risk minimization'' in drug discovery---how does one
minimize the chances of a small molecule failing as a drug candidate
and how do cheminformatics methods play a key role in these efforts.

@@

There is a considerable progress in open chemical and bioactivity databases like
ChEMBL and NCBI-PubChem, which are both freely available. Still, the integration
of chemical databases with one other remains a challenge, not only due to the
normalization problems in the chemical and bioactivity landscape, but also due
to sheer data volume. As mentioned earlier, we have to be able to
search in complex data, for example multi-labelled molecular graphs, where
graph labels can be different. Still, we need to be able to capture
molecular graphs in a risk-bounded machine encoded way.

Chemical graphs can differ based on the manner in which the structures are
drawn. Also chemical behaviors can result in multiple encodings of ``the same'' molecule and its
tautomeric states, different protonation states, or many
variations of implicit chemical expert knowledge. Checking whether two graphs represent the same chemical is an expensive
operation (the next section will discuss this in-depth); thus, a representation
of the structure that is unique and invariant with respect to how the structure
is drawn is needed for identification. An early effort in this direction was the
Simplified Molecular Input Line Specification (SMILES). However, several
different SMILES implementations became available, with no universal standard of interpretation. After years of research and various
commercial products, there is today no single public standard for a
line-notation of molecular labeled graphs, which is available to a broad
audience (academic and commercial). %The very same problem does also exist for
%bioassay ontologies and the lack of options to normalize the property spaces of
%chemical compounds, e.g. biological assays or clinical data outcome
%(biobanking).

With each database using their own algorithm for molecular encoding, the automated exchange of data between different databases was not facilitated. As more
and more data became available online, structure-based identification became a
pressing need, and the IUPAC International Chemical Identifier (InChI) code was
developed. The InChI is a non-proprietary, structured textual identifier for
chemical entities [1]. InChI identifiers are not intended to be read and
understood by humans, but are useful for computational matching of chemical
entities. For rapid database lookups, the InChIKey is a hashed key for the
InChI, which has an invariant length of 14 characters. Figure 1 illustrates the
SMILES, InChI and InChIKey for lipoic acid.

%Figure 1: The standard chemical graph illustration together with the SMILES,
%InChI and InChIKey codes for lipoic acid. Note that the SMILES is much more
%human-readable than the others.

The InChI is now widely used for matching identical chemical structures, but it
is still limited. For example, it cannot differentiate between
certain types of stereoisomers: Figure 2 illustrates two stereoisomers for which
the generated InChI is the same.

%Figure 2: The generated InChI is the same for the stereoisomers cisplatin and
%transplatin.

InChI or other identity-mapping algorithms allow for exact searching (within the
limitations of the algorithms), but two other requirements for scientific
discovery based on chemical databases are \emph{substructure searching} and \emph{similarity
searching}. In substructure searching, the database is searched for a specified
wholly contained part of the structures in the database; in similarity
searching, the database is searched for structures similar to a provided search
structure. Chemical search packages are often implemented and optimised for a
given database technology, for example, the OrChem package is an open source
chemical search package for the Oracle database \cite{rijnbeek2009}.

Graph substructure matching is a variant of \emph{graph isomorphism detection},which is
well known to be computationally intractable [3]. To execute a graph isomorphism
search across a full chemical database of thousands or even millions of
structures is completely infeasible. For that reason, complex heuristics must be
used to drastically filter the candidate search space. Structure-based
\emph{fingerprints} are commonly used for this purpose. Fingerprints encode
characteristic features of a given chemical structure in a boolean array, or
bitmap. The fingerprint is created by an algorithm which generates bitmap
patterns for each atom in the structure; then each atom and its nearest
neighbours, including the bonds between them; then each group of atoms connected
by paths of increasing lengths up to the maximum implemented for the algorithm
(commonly around eight). The bitmap patterns are added together into the final
fingerprint. If all bits in a query fingerprint are also present in the target
fingerprint of a stored database structure, only then is the structure subjected
to the computationally expensive subgraph matching algorithm. Bit operations are
very fast and independent of the number of atoms in a structure due to the fixed
length of the fingerprint.

Fingerprints also facilitate the rapid calculation of quantitative
structure-based similarity measures. An example measure of similarity is the
\emph{Tanimoto coefficient}, calculated as the ratio $T(a,b) = \frac{c}{(a+b−c)}$,
where c is the count of bits on in the same position in both the two
fingerprints, a is the count of bits on in the first structure, and b is the
count of bits on in the second structure. The Tanimoto coefficient varies in the
range 0.0 -- 1.0, with a score of 1.0 indicating that the two structures are very
similar (i.e. their fingerprints are the same). Structure-based similarity is of
limited use in grouping of chemicals based on non-structural features, such as
shared bioactivity profiles. For this purpose, a large-scale and flexible
classification structure is needed for chemicals and their linked biological
information. One such answer to this need is provided in the form of \emph{ontologies}.
An ontology is a formal specification of entities and their relationships in a
domain of interest. It is structured around an underlying logical representation
such as Description Logic [baaderdl2007]. The logic-based representation allows
automated reasoning to derive new knowledge (inferences) from the knowledge
encoded, making the representation both compact and powerful.
Ontologies are already in widespread use across the biomedical domain. A
similarity measure based on the information encoded in ontologies is termed
\emph{semantic similarity}, and such measures have already been applied to enhance
chemical classification in several problem areas [couto2010]. Many open
questions still remain about the best way to combine the information encoded in
chemical structures with chemical ontologies \cite{hastingsowled2010}.

@@

Enumerating molecules is a combinatorial problem that has fascinated chemists,
computer scientists and mathematicians alike for more than a century. It is one
important risk-reduction element in still being able to encode closely-related
molecules and for keeping the space complexity of databases under (algorithmic)
control. Indeed, while trying to solve the problem of counting the isomers of
paraffin structures or counting substituted aromatic compounds, fundamental
principles in graph theory and combinatorics were developed by Cayley, Polya and
others. Even terms that are widely-used nowadays like graph and tree were
originally coined in a chemistry context (Chapters 1 and 8 in [1]). More than
four decades ago, Joshua Lederberg, Carl Djerassi, Edward Feigenbaum and others
from Stanford University started developing codes to explicitly enumerate
molecules. Their project, named DENDRAL, was funded by NASA with the goal of
developing an automated system to elucidate compounds from spectral data. While
studying this challenging problem, important concepts in computer science were
developed. DENDRAL is generally quoted in artificial intelligence textbooks as a
pioneer project that produced the first expert system.

Enumerating molecules is not only an interesting academic exercise but has
practical utility as well. The foremost application of enumeration is \emph{structure
elucidation} of natural compounds such as metabolites, geochemicals, and
petrochemicals. Ideally, the wishful bench chemist collects experimental data
from an unknown compound, the data is fed into a code, and the resulting
structure is returned. Although that streamlined process may not yield a unique
solution, there are commercial software products such as MOLGEN
(http://molgen.de) that can, for instance, list all structures matching a given
molecular formula, or mass spectrometry (MS) data. Another important application
of enumeration is \emph{molecular design}. Here the problem is to design compounds
(drugs, for instance) that optimize some physical, chemical, or biological
property or activity. Although less prolific and more recent than structure
elucidation, molecular design has introduced some novel solutions to molecular
enumeration. Finally, with the advent of combinatorial chemistry, molecular
enumeration has taken a central role as it allows computational chemists to
construct virtual libraries, test hypotheses, and provides guidance in the
design of optimal combinatorial experiments.

The major difficulty with enumeration is that the \emph{in silico} representation of
molecules, the so-called molecular graphs, are labeled object (i.e. labeled
graphs), while the atoms of a molecule are of course not uniquely labeled. The
mathematical concept to tackle this problem is to consider orbits of labeled
molecular graphs under the operation of the symmetric group. When enumerating
molecular structures, one has thus to solve the so--called isomorphism problem.
The good news is that isomorphism can be solved in polynomial time for molecular
graphs due to the fact that molecules belong to a restricted class of graphs
known as \emph{bounded valence graphs}; however, that same restriction prevents
applying to a molecular graph the polynomial delay enumeration results obtained
for general graphs (Chapter 8 in [1]). While the complexity of molecular graph
enumeration remains an open problem it has nonetheless been shown that molecules
can be sampled efficiently [2]. Sampling procedures based on the Metropolis or
the Genetic Algorithms have been developed and used to elucidate natural
compounds from NMR data.

The underlying computational complexity and potential intractability of
molecular enumeration may not be a bottleneck after all, as computational
chemists have developed and successfully used enumeration tools to generate
large chemical libraries like GDC-13 comprising almost a trillion chemical
structures\footnote{http://www.dcb-server.unibe.ch/groups/reymond/gdb/home.html}. Yet,
the current enumeration software products do not generally produce stereoisomers
[3] or tautomers [4]; these require specific enumeration procedures, which are
still the subject of investigations by the cheminformatics community.

GDB-13 along with other databases such as PubChem (NCBI/NIH) or ZINC (UCSF) are
valuable resources to search for new drugs. For instance the well known
ligand-based design approach is making use of these databases along with
structure similarity tools to discover new compounds binding to biological
targets (protein or nucleic acid). Compounds retrieved through ligand-based
design, however, must already be present in databases, which is a limitation of
the technique. To palliate that limitation, algorithms have recently been
developed to enumerate all compounds in the chemical space matching circular [5]
or path fingerprints [6], these fingerprints being popular tools to measure
similarity. Enumerating molecules from fingerprints, or more generally
enumerating molecules from structure-activity molecular descriptors, is still
very much under development. At stake is the design of new chemicals or drugs
that have not yet been synthesized.

Another area where enumeration plays a role is in the \emph{generation of chemical
reaction networks}. The problem here consists of enumerating all the possible
compounds that can be produced by applying reaction rules to a set of initial
molecules. By reversing the reaction rules one can also find sets of starting
compounds necessary for producing a given target. This latter process is known
in chemistry as \emph{retrosynthesis}, ever since the works of E.J. Corey, who received the
Nobel Prize in Chemistry in 1990. Designing new drugs or chemicals,
understanding the kinetics of combustion and petroleum refining, studying the
dynamics of metabolic networks, applying metabolic engineering and synthetic
biology to produce heterologous compounds in microorganisms, all are
applications that require the enumeration of reaction networks. As reviewed in Chapter
11 in [1] several network enumeration techniques have been developed. However,
these techniques generally suffer from a combinatorial explosion of intermediate
compounds being produced. One way to limit the number of compounds generated is
to simulate the dynamics of the network while it is being constructed and remove
compounds of low concentration. Following that idea, methods have been
developed based on the Gillespie Stochastic Simulation Algorithm (SSA) to
compute on-the-fly species concentrations. Chemical reaction network enumeration
and sampling is an active field of research, particularly in the context of
metabolism, either to study biodegradation, or to propose metabolic engineering
strategies to biosynthesize compounds of commercial interest. The difficulty
with metabolic network design is that in addition to network generation based on
reactions, one also needs to verify that there are enzymatic sequences enabling
reactions catalyses. That additional task requires encompassing both chemical
and sequence information and the development of tools that are at the interface
between chemoinformatics and bioinformatics.
